\Kotac{} is an internal Domain Specific Language for the aggregate programming in Kotlin, currently under development at the address \textit{github.com/kklorenzotesta/KotAC}. Its designed to be fully cross-platform with easy to use idiomatic APIs. Like \Scafi{} the internal DSL offers an easy integration with existing code, and the API has the same semantics of the field calculus. It is also designed to be usable for Android applications. The repository has been integrated with the Travis \cite{Travis} continous integration to validate the pushed builds. The library is built with the Gradle \cite{Gradle} build tool, the standard build tool for Kotlin projects, and linted with the ktlint \cite{ktlint} plugin.

\section{Explicit fields}
Unlike \Scafi{}, in which the fields are implicitly represented in the notion of computation against a neighbour, \Kotac{} explicitly represents the field values with the following interface:
\begin{lstlisting}[language={kotac}]
interface Field<out T> {
    fun <R> map(f: (T) -> R): Field<R>
    fun <R> fold(initial: R, operation: (acc: R, T) -> R): R
}
\end{lstlisting}
$\mathtt{Field}$ is an interface with a covariant\footnote{A class or interface $\mathtt{Foo<T>}$ is covariant in $\mathtt{T}$ if, for each class $\mathtt{B}$ subclass of $\mathtt{A}$, $\mathtt{Foo<B>}$ is subclass of $\mathtt{Foo<A>}$} type parameter $\mathtt{T}$, representing the type of the values to which each device in the field is mapped to. A field exposes only two high level functions for its manipulation. The $\mathtt{map}$ function applies the function $\mathtt{f}$, from $\mathtt{T}$ to $\mathtt{R}$, to each value of the field, obtaining a new field $\mathtt{Field<R>}$. The $\mathtt{fold}$ function allow to reduce a field to a regular local value. Like the \Scafi{} $\mathtt{foldhood}$ the $\mathtt{fold}$ function repeatedly combines the values of the field into an accomulator of type $\mathtt{R}$ by applying the given $\mathtt{operation}$ to each value starting with the value of $\mathtt{initial}$.

Based on this basic operators the \Kotac{} standard library and the users can define new extension functions for the most common operations on fields. For example the built-in operator $\mathtt{anyhood}(\fvalue)$, which evaluates to true if any value of the field is true, can be defined as:
\begin{lstlisting}[language={kotac}]
fun Field<Boolean>.any(): Boolean = this.fold(true) { acc, v -> acc || v  }
\end{lstlisting}

A concrete implementation of field has been defined only for internal use. The implementation is backed by an immutable $\mathtt{Map}$ associating $\mathtt{DeviceID}$s to the values. The implementation is not part of the public API, disallowing arbitrary manipulations of fields data that may alter the semantics.

$\mathtt{DeviceID}$ is an empty interface representing an identifier for a device. Two implementations are provided, one based on integer identifiers and one based on string identifiers.

\section{Basic operators}

The API for the field calculus operators are defined inside the interface $\mathtt{AggregateContext}$:
\begin{lstlisting}[language={kotac}]
interface AggregateContext {
    fun <T> share(initial: T, f: (Field<T>) -> T): T
    fun <K, T> align(key: K, proc: (K) -> T): T
    fun <T> nbr(f: () -> T): Field<T>
    fun <T> rep(initial: T, f: (T) -> T): T
}
\end{lstlisting}
The $\mathtt{share}$ function implements the share operator of the field calculus, which at each execution applies the neighbouring field of the last computed value to the function $\mathtt{f}$, starting with the $\mathtt{initial}$ value for the executing device. The function $\mathtt{align}$ executes the function $\mathtt{proc}$ aligning to the devices with the same value of $\mathtt{key}$. The function $\mathtt{nbr}$ implements the nbr operator of the field calculus, running the function $\mathtt{f}$ for the executing device and returning a neighbouring field with the last values. The function $\mathtt{rep}$ implements the rep operator of the field calculus, which, starting from $\mathtt{initial}$ applies the last computed value to $\mathtt{f}$.

Aggregate programs can then be written as extension function of $\mathtt{AggregateContext}$, allowing to call the operators just like keywords. For example the following code:
\begin{lstlisting}[language={kotac}]
fun AggregateContext.rounds(): Int = rep(0) { x -> x + 1 }
\end{lstlisting}
defines an aggregate program called $\mathtt{rounds}$ which counts the number of rounds that has been executed, if called at every round on the same $\mathtt{AggregateContext}$.

\section{Internal implementation}
The $\mathtt{ConcreteAggregateContext}$ class implements the $\mathtt{AggregateContext}$ interface providing the implementation of all the operators. 

Internally the $\mathtt{Trace}$ interface represents an immutable point in the execution that will be use for the alignments,  it doesn't defines any methods but its used only for equality. The interface $\mathtt{Stack}$ is defined as:
\begin{lstlisting}[language={kotac}]
interface Stack {
    fun enter(align: Any)
    fun exit()
    val trace: Trace
}
\end{lstlisting}
It offers methods to manage the creating of $\mathtt{Trace}$s. The $\mathtt{enter}$ and $\mathtt{exit}$ methods are called by the operators to update che current $\mathtt{Trace}$, the method $\mathtt{trace}$ returns then a new instance of $\mathtt{Trace}$ based on the previous invocations of $\mathtt{enter}$ and $\mathtt{exit}$.

The internal state of a $\mathtt{ConcreteAggregateContext}$ is kept in the data structures $\mathtt{ValueTree}$ and $\mathtt{ValueEnvironment}$. A $\mathtt{ValueTree}$ is an interface defined as:
\begin{lstlisting}[language={kotac}]
interface ValueTree {
    fun get(trace: Trace): Any?
    fun put(trace: Trace, value: Any?)
}
\end{lstlisting}
The inferface describes a basic mutable $\mathtt{Map}$ from $\mathtt{Trace}$s to values of any type. Since values are absent in non aligned devices the returned values are nullable.
A $\mathtt{ValueEnvironment}$ is an interface defined as:
\begin{lstlisting}[language={kotac}]
interface ValueEnvironment {
    fun getTree(device: DeviceID): ValueTree
    fun putTree(device: DeviceID, tree: ValueTree)
}
\end{lstlisting}
The interface describes a basic mutable $\mathtt{Map}$ from $\mathtt{DeviceID}$s to $\mathtt{ValueTree}$s.
Both $\mathtt{ValueTree}$ and $\mathtt{ValueEnvironment}$ are not designed to be accessed by application developers, but instead to be managed by the basic operators. Therefore their API will be modified according to the needs of the internals.

The $\mathtt{ConcreteAggregateContext}$ contains then a field named $\mathtt{localTree}$ which is a $\mathtt{ValueTree}$ in which are stored the data for the local interactions, i.e. the rep operator. Another field named $\mathtt{neighboringTree}$ is a $\mathtt{ValueTree}$ in which are stored the data for the neighbouring interactions, i.e. the share and nbr operators, this tree will be sent to the neighbours. Finally a field named $\mathtt{neighboringEnvironment}$ is a $\mathtt{ValueEnvironment}$ holding all the last $\mathtt{ValueTree}$s sent by the neighbours.

\section{\FKotac}

\subsection{\FKotac{} syntax}

\begin{figure}[t]
\centering
\centerline{$
\begin{array}{l@{\hspace{1mm}}c@{\hspace{1mm}}l@{\hspace{-8mm}}r}
        \PROGRAM & \BNFcce & \overline{\FUNCTION} \; \aexp
                                                                                                                                                                                                        &   {\footnotesize \mbox{program}} \\[3pt]
        \FUNCTION & \BNFcce &  \SFUNCTION \, \BNFmid \, \AFUNCTION
                                                                                                                                                                                                        &   {\footnotesize \mbox{function declaration}} \\[3pt]
        \hline \\[-6pt]
        \SFUNCTION & \BNFcce &  \funK \; \sfname (\overline{\xname}) \; \{ \sexp \}
                                                                                                                                                                                                        &   {\footnotesize \mbox{standard function declaration}} \\[3pt]
        \sexp & \BNFcce & \xname \, \BNFmid \, \sanyvalue \, \BNFmid \, (\overline{\xname}) \; \ftoSym \{ \sexp \} \, \BNFmid \, \sexp(\overline{\sexp})
                                                                                                                                                                                                        &   {\footnotesize \mbox{standard expression}} \\[3pt]
                \sanyvalue & \BNFcce &  \lvalue \; \BNFmid \; \sfunvalue
                &{ \mbox{\footnotesize standard value}}
                \\[3pt]
                \lvalue & \BNFcce &  \dcOf{\dc}{\overline\lvalue}
                &{ \mbox{\footnotesize local data value}}
                \\[3pt]
        \sfunvalue & \BNFcce & \sbname \; \BNFmid \; \sfname \; \BNFmid \; (\overline{\xname}) \; \ftoSym \{ \sexp \}
                                                                                                                                                                                                        &   {\footnotesize\mbox{standard function value}} \\
        \hline \\[-6pt]
        \AFUNCTION & \BNFcce &  \funK \; \afname (\overline{\xname}) \; \{ \alignK{\{\aexp\}} \}
                                                                                                                                                                                                        &   {\footnotesize \mbox{aggregate function declaration}} \\[3pt]
        \aexp & \BNFcce & \xname \, \BNFmid \, \anyvalue  \BNFmid \; (\overline{\xname}) \; \toSym{\name}{\{\alignK\{\aexp\}\}} \, \BNFmid \, \aexp(\overline{\aexp}) \, \BNFmid \,  &   {\footnotesize \mbox{aggregate expression}} \\
        && \repK(\aexp)\{ (\xname) \ftoSym \aexp \} \, \BNFmid \, \nbrK\{\aexp\} \, \BNFmid \, \shareK(\aexp)\{(\xname) \ftoSym \aexp \} \\[3pt]
                \anyvalue & \BNFcce &  \lvalue \; \BNFmid \; \funvalue \; \BNFmid \; \fvalue
                &{ \mbox{\footnotesize value}}
                \\[3pt]
              \funvalue & \BNFcce &  \sfunvalue \; \BNFmid \; \afunvalue
                &{ \mbox{\footnotesize function value}}
                \\[3pt]
                \fvalue & \BNFcce &  \envmap{\overline\deviceId}{\overline\lvalue}
                &{ \mbox{\footnotesize neighbouring field value}} \\
        \afunvalue & \BNFcce & \abname \; \BNFmid \; \afname \; \BNFmid \; (\overline{\xname}) \; \toSym{\name}{\{\alignK\{\aexp\}\}}
                                                                                                                                                                                                        &   {\footnotesize\mbox{aggregate function value}} \\
\end{array}
$
}
\caption{Syntax of \FKotac{}}
\label{fig:fkotacsyntax}
\end{figure}

In order to formalize the semantic of \Kotac{} a subset of it has been considered called \textit{Featherweight \Kotac{} (\FKotac{})}. The syntax of \FKotac is shown in figure \ref{fig:fkotacsyntax}, it focus on the aggregate constructs ignoring the Kotlin specific features. Unlike \FSCAFI, which considers only aggregate functions (which require alignment), in \FKotac has been considered also standard function, i.e. kotlin functions that don't require alignment and don't allow to call aggregate functions and aggregate constructs in their body. For simplicity data values of function values and fields of function values are not allowed. Also standard function values cannot be passed to aggregate functions and viceversa.

A program $\PROGRAM$ is a sequence of function declarations $\overline{\FUNCTION}$, each of which can be either a standard function $\SFUNCTION$ or an aggregate function $\AFUNCTION$, followed by a main aggregate expression $\aexp$. A standard function declaration $\SFUNCTION$ defines a (possibly recursive) standard function named $\sfname$ with a sequence of formal parameters $\overline{\xname}$ and a body standard expression $\sexp$. An aggregate function declaration $\AFUNCTION$ defines a (possibly recursive) aggregate function named $\afname$, with a sequence of formal parameters $\overline{\xname}$ and a body aggregate expression $\aexp$. 

A standard expression $\sexp$ can be a variable $\xname$, a standard value $\svalue$, an anonymous function $(\overline{\xname}) \; \ftoSym \{ \sexp \}$  where $\overline{\xname}$ are variable names for the formal parameters and $\sexp$ is the body, or a function application $\sexp(\overline{\sexp})$ where $\sexp$ evaluates to a function $\sfunvalue$ and $\overline{\sexp}$ are the arguments. An aggregate expression $\aexp$ can be a variable $\xname$, a value $\anyvalue$, an anonymous function $(\overline{\xname}) \; \toSym{\name}{\{\alignK\{\aexp\}\}}$ where $\overline{\xname}$ are variable names for the formal parameters, $\aexp$ is the body of the function and $\name$ is a \textit{tag} that uniquely identifies the function and doesn't occours in the source program, a function application $\aexp(\overline{\aexp})$ where $\aexp$ evaluates to a function $\afunvalue$ or $\sfunvalue$ and $\overline{\aexp}$ are the arguments, a $\mathtt{rep}$-expression $\repK(\aexp)\{ \aexp \}$ modelling time evolution, a $\mathtt{nbr}$-expression $\nbrK\{\aexp\}$ modelling neighbourhood interaction or a $\mathtt{share}$-expression $\shareK(\aexp)\{(\xname) \ftoSym \aexp \}$ modelling time evolution based on a neighbouring field.

A standard value can be either a local data value $\lvalue$ or a standard function value $\sfunvalue$. A local data value is a data value $\dcOf{\dc}{\overline\lvalue}$, in which $\dc$ is a data constructor and $\overline{\lvalue}$ are local data value arguments. A value can be either a local data value $\lvalue$, a function value $\funvalue$ or a neighbouring field value $\fvalue$. Neighbouring field values doesn't appear in the source code but can only be computed dynamically, usally by built-in operators like $\mathtt{nbr}$ or $\mathtt{share}$. A function value can be either a standard function value $\sfunvalue$ or an aggregate function value $\afunvalue$.

A standard function value $\sfunvalue$ can be a standard built-in function $\sbname$ (e.g. a temperature sensor or a pure mathematical or logical function), a declare standard function $\sfname$ or a closed anonymous standard function value $(\overline{\xname}) \; \ftoSym \{ \sexp \}$. An aggregate function value $\afunvalue$ can be an aggregate built-in function $\abname$ (e.g. nbrsense, which returns a field of the distances from the neighbours, or maxhood, which returns the maximum value in the argument field), a declared aggregate function $\afname$ or a closed anonymous aggregate function value $ (\overline{\xname}) \; \toSym{\name}{\{\alignK\{\aexp\}\}}$.

Both aggregate branching expression and standard if expression are not part of the syntax since they can be implemented as syntactic sugar, they will therefore not be considered in the semantic and typing rules.

\subsection{\FKotac{} semantics}

\begin{figure}[pt]{
 \scalebox{0.85}{
 $\begin{array}{l}
 \textbf{Value-trees and value-tree environments:}\\
\begin{array}{lcl@{\hspace{6.8cm}}r}
%
\vtree & \BNFcce &  \mkvt{\anyvalue}{\overline{\vtree}}    &   {\footnotesize \mbox{value-tree}} \\
\Trees & \BNFcce & \envmap{\overline{\deviceId}}{\overline{\vtree}}   &   {\footnotesize \mbox{value-tree environment}}
%
\end{array}\\[10pt]
\hline\\[-8pt]
%%%%  AUX
\textbf{Auxiliary functions:}\\
\begin{array}{l}
\begin{array}{l@{\hspace{1cm}}l}
%
\vrootOf{\mkvt{\anyvalue}{\overline{\vtree}}}  =   \anyvalue
\\
\vroot^\funvalue(\mkvt{\anyvalue}{\overline{\vtree}})  = \mkvt{\anyvalue}{} \text{ if } \funvalue \text{ standard else } \mkvt{\anyvalue}{\overline{\vtree}}
\\
%
\piIof{i}{\mkvt{\anyvalue}{\vtree_1,\ldots,\vtree_n}}  =   \vtree_i
\quad \mbox{if} \; 1\le i \le n
\\
\piBof{\funvalue}{\mkvt{\anyvalue}{\vtree_0,\ldots,\vtree_{n+1}}}  =   \vtree_{n+1}
\quad \mbox{if} \; \nameOf(\vrootOf{\vtree_{0}}) = \nameOf(\funvalue) \text{ and } \funvalue \text{ not standard}
\\
 \piBof{\funvalue}{\vtree}  =   \emptyseq \quad \mbox{otherwise}
\\  
\end{array}
\\
\mbox{For } \auxNAME\in\rho,\piI{i},\piB{\funvalue}:
\quad 
\left\{\begin{array}{lcll}
 \aux{\envmap{\deviceId}{\vtree}, \Trees}  & =  & \envmap{\deviceId}{\aux{\vtree}}, \aux{\Trees} & \quad \mbox{if} \; \aux{\vtree} \not=\emptyseq  
\\
\aux{\envmap{\deviceId}{\vtree}, \Trees}  & =   & \aux{\Trees} & \quad \mbox{if} \; \aux{\vtree}=\emptyseq  
\\
\aux{\emptyseq}  & =  &  \emptyseq
\end{array}\right.   
\\
\begin{array}{ll}
\nameOf(\fname) = \fname \qquad \nameOf(\bname) = \bname
&
\nameOf((\overline{\xname}) \; \toSym{\name}{\{\alignK\{\e\}\}}) = \name
\\
\args{\fname} = \overline{\xname} \quad \mbox{if } \, \funK \; \fname (\overline{\xname}) \; \{ \alignK{\{\e\}} \}
&
\args{\fname} = \overline{\xname} \quad \mbox{if } \, \funK \; \fname (\overline{\xname}) \; \{ \e \}
\\
\args{(\overline{\xname}) \; \toSym{\name}{\{\alignK\{\e\}\}}} = \overline{\xname}
&
\args{(\overline{\xname}) \; \ftoSym \{ \e \}} = \overline{\xname}
\\
\body{\fname} = \e  \quad \mbox{if } \, \funK \; \fname (\overline{\xname}) \; \{ \alignK{\{\e\}} \}
&
\body{\fname} = \e  \quad \mbox{if } \,  \funK \; \fname (\overline{\xname}) \; \{ \e \}
\\
\body{(\overline{\xname}) \; \toSym{\name}{\{\alignK\{\e\}\}}} = \e
&
\body{(\overline{\xname}) \; \ftoSym \{ \e \}} = \e
\end{array}
\\
\begin{array}{l@{\hspace{0.4cm}}l}
		\fvalue_0[\fvalue_1] = \fvalue_2 \; \text{ where } \fvalue_2(\deviceId) = \left\lbrace \begin{array}{ll}
			\fvalue_1(\deviceId) & \text{if } \deviceId \in \domof{\fvalue_1} \\
			\fvalue_0(\deviceId) & \text{otherwise}
		\end{array} \right. \\
\end{array}
\end{array}\\
\hline\\[-10pt]
\textbf{Syntactic shorthands:}\\
\begin{array}{l@{\hspace{5pt}}l@{\hspace{5pt}}l}
\bsopsem{\deviceId}{\piIofOv{\Trees}}{\senstate}{\overline{\e}}{\overline{\vtree}}
&
  \textrm{where~~} |\overline{\e}|=n
&
  \textrm{for~~}
  \bsopsem{\deviceId}{\piIof{1}{\Trees}}{\senstate}{\e_1}{\vtree_1}
    \cdots
    \bsopsem{\deviceId}{\piIof{n}{\Trees}}{\senstate}{\e_n}{\vtree_n} \!\!\!\!\!\!\!\!\!\!\!\! \\
\vrootOf{\overline{\vtree}}
&
  \textrm{where~~} |\overline{\vtree}|=n
  & \textrm{for~~}
\vrootOf{\vtree_1},\ldots,\vrootOf{\vtree_n}\\
\substitution{\overline{\xname}}{\vrootOf{\overline{\vtree}}}
&   \textrm{where~~} |\overline{\xname}|=n
  &
  \textrm{for~~}
\substitution{\xname_1}{\vrootOf{\vtree_1}}~\ldots\quad\substitution{\xname_n}{\vrootOf{\vtree_n}}
\end{array}\\
\hline\\[-10pt]
%%%  EVALUATION RULES
\textbf{Rules for expression evaluation:} \hspace{4.4cm} %\hfill
%\vspace{-0.2cm}
  \boxed{\bsopsem{\deviceId}{\Trees}{\senstate}{\e}{\vtree}}
\skiptransition%[-5pt]
\begin{array}{c}
%\vspace{-0.1cm}
\surfaceTyping{E-LOC}{
	\e = \lvalue \text{ or } \funvalue
}{
	\bsopsem{\deviceId}{\Trees}{\senstate}{\e}{\mkvt{\e}{}}
}
\qquad\qquad
\surfaceTyping{E-FLD}{\qquad \fvalue' = \proj{\fvalue}{\domof{\Trees}\cup\{\deviceId\}}}{
\bsopsem{\deviceId}{\Trees}{\senstate}{\fvalue}{\mkvt{\fvalue'}{}}
}
\skiptransition\\[-6pt]
\surfaceTyping{E-B-APP}{  \quad
\begin{array}{c}
  \bsopsem{\deviceId}{\piIofOv{\Trees}}{\senstate}{\overline{\e},\e}{\overline{\vtree},\vtree}
  \qquad \bname = \vrootOf{\vtree}
  \qquad \vtree'=\builtinop{\bname}{\correction{\deviceId}}{\piBof{\bname}{\Trees},\senstate}(\vrootOf{\overline{\vtree}})
\end{array}
 }{
\bsopsem{\deviceId}{\Trees}{\senstate}{\e(\overline{\e})}{\mkvt{\vrootOf{\vtree'}}{\vtree,\overline{\vtree},\vroot^\bname(\vtree')}}
}
%
\skiptransition\\[-6pt]
%
\surfaceTyping{E-D-APP}{ \quad
\begin{array}{c}
  \bsopsem{\deviceId}{\piIofOv{\Trees}}{\senstate}{\overline{\e},\e}{\overline{\vtree},\vtree} \qquad 
  \funvalue=\vrootOf{\vtree} \mbox{ not a built-in} \qquad
\\
  \bsopsem{\deviceId}{\piBof{\funvalue}{\Trees}}{\senstate}{\applySubstitution{\body{\funvalue}}{\substitution{\args{\funvalue}}{\vrootOf{\overline{\vtree}}}}}{\vtree'}
\end{array}
}{
\bsopsem{\deviceId}{\Trees}{\senstate}{\e(\overline{\e})}{\mkvt{\vrootOf{\vtree'}}{\vtree,\overline{\vtree},\vroot^\funvalue(\vtree')}}
}
%
\skiptransition\\[-5pt]
\surfaceTyping{E-NBR}{
         \qquad
     \Trees_1=\piIof{1}{\Trees} \qquad
     \bsopsem{\deviceId}{\Trees_1}{\senstate}{\e}{\vtree_1}
\qquad
 \fvalue=\mapupdate{\vrootOf{\Trees_1}}{\envmap{\deviceId}{\vrootOf{\vtree_1}}}
 }{
\bsopsem{\deviceId}{\Trees}{\senstate}{\nbrK\{\e\}}{\mkvt{\fvalue}{\vtree_1}}
}
\skiptransition\\[-6pt]
\surfaceTyping{E-REP}{
        \quad
        \begin{array}{l}
     \bsopsem{\deviceId}{\piIof{1}{\Trees}}{\senstate}{\e_1}{\vtree_1} \\
     \bsopsem{\deviceId}{\piIof{2}{\Trees}}{\senstate}{\applySubstitution{\e_2}{\substitution{\xname}{\lvalue_0}}}{\vtree_2}~~
        \end{array}
        \quad
        \lvalue_0 \! = \!\left\{\begin{array}{ll}
                             \vrootOf{\piIof{2}{\Trees}}(\deviceId) & \mbox{if} \;  \deviceId \in \domof{\Trees} \\
                             \vrootOf{\vtree_{1}} & \mbox{otherwise}
                           \end{array}\right.
 }{
\bsopsem{\deviceId}{\Trees}{\senstate}{\repK(\e_1)\{(\xname) \; \ftoSym \; \e_2\}}{\mkvt{\vrootOf{\vtree_{2}}}{\vtree_1,\vtree_2}}
}
\skiptransition\\[-4pt]
\surfaceTyping{E-SHARE}{ \qquad
	\begin{array}{l@{\hspace{0.5em}}l}
     \bsopsem{\deviceId}{\piIof{1}{\Trees}}{{\senstate}}{\e_1}{\vtree_1} & \fvalue' = \vrootOf{\piIof{2}{\Trees}} 
      \qquad \qquad \fvalue = (\envmap{\deviceId}{\vrootOf{\vtree_1}})[\fvalue']
     \\
     \bsopsem{\deviceId}{\piIof{2}{\Trees}}{{\senstate}}{\applySubstitution{\e_2}{\substitution{\xname}{\fvalue}}}{\vtree_2} %& \fvalue = (\envmap{\deviceId}{\vrootOf{\vtree_1}})[\fvalue']
	\end{array}
	\!\!\!\!
 }{
	\bsopsem{\deviceId}{\Trees}{\senstate}{\shareK(\e_1)\{(\xname) \; \ftoSym \; \e_2\}}{\mkvt{\vrootOf{\vtree_{2}}}{\vtree_1,\vtree_2}}
}
\end{array}
\end{array}$
}
}
 \caption{Big-step operational semantics of \FKotac{}.} \label{fig:kotacSemantics}
\end{figure}

Figure \ref{fig:kotacSemantics} shows the operational semantics of the syntax. It follows the same notation of the field calculus semantics and uses the same auxiliary functions. Rules \ruleNameSize{[E-LOC]} e \ruleNameSize{[E-FLD]} define the evaluation of local data values, function values and neighbouring values, both producing a value-tree with no subtrees, with a domain restriction to the aligned devices in case of neighbouring fields. Rules \ruleNameSize{[E-B-APP]} and \ruleNameSize{[E-D-APP]} model the application of aggregate or standard built-in functions and aggregate or standard user defined (or anonymous) functions, the former based on a function different for each built-in and the latter by executing the body of the function. In the case of standard functions the values are computed with an empty value-tree environement. Rule \ruleNameSize{[E-NBR]} models the evaluation of $\mathtt{nbr}$-expressions, by extracting from the value-tree environment the neighbouring values to build a neighbouring field, and computing the expressions $\e$ for the new value of the executing device. Rule \ruleNameSize{[E-REP]} models the evauation of $\mathtt{rep}$-expressions, by extracting from the value-tree environemtn the last computed value for the new evaluation of $\e$. Rule \ruleNameSize{[E-SHARE]} models the evaluation of $\mathtt{share}$-expressions, by constructing a field of the last computing values as the argument $\xname$ for the evaluation of $\e_2$ and using $\e_1$ as the initial value of the executing device.

It's important to note that a standard function $\sfname$ can be called for example with an aggregate anonymous functions $\name$ as parameters according to the syntax and semantics. If the function $\name$ is then called inside the body of $\sfname$ it will not have the expected aligned behavior. In order to avoid this situation typing rules must enforce that only standard values can be passed as arguments of standard functions (either declared or anonymous).

\begin{figure}[!t]{
 $\begin{array}{l}
 %%%  SYNTAX
 \textbf{System configurations and action labels:}\\
\begin{array}{lcl@{\hspace{7.5cm}}r}
%
\Field & \BNFcce &  \envmap{\overline\deviceId}{\overline\Trees}    &   {\footnotesize \mbox{status field}} \\
\Topo & \BNFcce &  \envmap{\overline\deviceId}{\overline\devset}    &   {\footnotesize \mbox{topology}} \\
\Sens & \BNFcce &  \envmap{\overline\deviceId}{\overline\senstate}    &   {\footnotesize \mbox{sensors-map}} \\
\Envi & \BNFcce &  \EnviS{\Topo}{\Sens}    &   {\footnotesize \mbox{environment}} \\
\Cfg & \BNFcce &  \SystS{\Envi}{\Field}    &   {\footnotesize \mbox{network configuration}} \\
\act & \BNFcce &  \deviceId \;\BNFmid\; \envact    &   {\footnotesize \mbox{action label}} \\
%
\end{array}\\
\hline\\[-8pt]
\textbf{Environment well-formedness:}\\
\begin{array}{l}
%
\wfn{\EnviS{\Topo}{\Sens}} \textrm{~~holds iff {$\domof{\Topo}=\domof{\Sens}$ and $\Topo(\deviceId) \subseteq \domof{\Sens}$ for all $\deviceId \in \domof{\Sens}$.}}
\\
\end{array}\\
\hline\\[-8pt]
%%%%%%%%%%
%%%  REDUCTION RULES
\textbf{Transition rules for network evolution:} \hfill
  \boxed{\nettran{\Cfg}{\act}{\Cfg}}
  \\[0.2cm]
\vspace{0.5cm}
\begin{array}{c}
\netopsemRule{N-FIR}
                 {\quad \Envi=\EnviS{\Topo}{\Sens}
                   \quad \Topo(\deviceId)= \overline\deviceId 
                  \quad {\bsopsem{\deviceId}{\filter(\Field)(\deviceId)}{\Sens(\deviceId)}{\emain}{\vtree}}
                  \quad
                 \Field_1=\envmap{\overline\deviceId}{\{\envmap{\deviceId}{\vtree}\}}}
                 {\nettran{\SystS{\Envi}{\Field}}{\deviceId}{\SystS{\Envi}{\mapupdate{\filter(\Field)}{\Field_1}}}%\nopsem{\EnviS{\Topo}{\Sens}}{\Field}{\deviceId}{\mapupdate{\Field}{\deviceId}{\vtree}}
                 }
\skiptransition
\netopsemRule{N-ENV}
                 {\qquad \wfn{\Envi'}\qquad \Envi'=\EnviS{\Topo}{\envmap{\overline\deviceId}{\overline\senstate}} \qquad
                  %\bsopsem{\senstate_1}{\emptyset}{\emain}{\vtree_1}\quad\cdots\quad\bsopsem{\senstate_n}{\emptyset}{\emain}{\vtree_n} \qquad
                  \Field_0=\envmap{\overline\deviceId}{\emptyset}
                 }
                 {\nettran{\SystS{\Envi}{\Field}}{\envact}{\SystS{\Envi'}{\mapupdate{\Field_0}{\Field}}}
                 }\\[-10pt]
\end{array}\\
%\textbf{Initial device state $\initialvtree$:}\\[0.2cm]
%\begin{array}{c}
%
%\qquad\qquad\netopsemRule{Init}
%                 {\senstate_0(\snsname)=\initialOf{\typeof{\snsname}}}
%                 {\bsopsem{\senstate_0}{\emptyset}{\emain}{\initialvtree}}
%
%\end{array}
%
\end{array}$
} %%%\capskip
 \caption{Small-step operational semantics for network evolution.} \label{fig:kotacnetworkSemantics}
\end{figure}

Figure \ref{fig:kotacnetworkSemantics} defines the operational semantics of whole networks. It uses the same syntax of the field calculus network semantics and the same rules. Like for the field calculus rule \ruleNameSize{[N-FIR]} models a computational round on device $\deviceId$ and rule \ruleNameSize{[N-ENV]} models an environement change to a new well-formed environment.

\subsection{\FKotac{} typing}

\begin{figure}[!t]{
 \scalebox{0.8}{
 $\begin{array}{l}
%%%  TYPES
\textbf{Types:}\\
\begin{array}{rcl@{\hspace{7.5cm}}r}
\type & \BNFcce &  \tvar  \; \BNFmid \;  \builtintype \; \BNFmid \;  (\overline\type) \rightarrow \type  \; \BNFmid \; \ftypeOf{\builtintype}       &   {\footnotesize \mbox{type}} \\
%
\builtintype & \BNFcce & \bvar\; \BNFmid \;  \bitype &   {\footnotesize \mbox{built-in type}} \\
%
\typescheme & \BNFcce &  \forall\overline{\tvar}\overline{\bvar}.\type      &   {\footnotesize \mbox{type scheme}} \\
%
\hline\\[-8pt]
\tctx & \BNFcce &  \aann \; \BNFmid \;  \sann &   {\footnotesize \mbox{context}} \\[8pt]
%
%\localenv & \BNFcce & \envS{\senstate}{\overline\vtree} &   {\footnotesize \mbox{local environment}} \\
\end{array}\\
\hline\\[-8pt]
%%%  TYPE RULES
\textbf{Expression typing:} 
  \hfill
  \boxed{\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv}{\e}{\type}}
\vspace{0.1cm}
  \\
\begin{array}{c}
\surfaceTyping{T-ATOS}{ \quad
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\e}{\builtintype}
}{
\expTypJudC{\sann}{\TStypEnv}{\TtypEnv}{\e}{\builtintype}
}
\qquad
\surfaceTyping{T-STOA}{ \quad
\expTypJudC{\sann}{\TStypEnv}{\TtypEnv}{\e}{\builtintype}
}{
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\e}{\builtintype}
}
\skiptransition
%
\nullsurfaceTyping{T-VAR}{
\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv,\xname:\type}{\xname}{\type}
}
\qquad
%
{
\surfaceTyping{T-DAT}{ \quad
\applySubstitution{\applySubstitution{\type'}{\substitution{\overline\tvar}{\overline\type''}}}{\substitution{\overline\bvar}{\overline\builtintype}} = (\overline{\type})\rightarrow\type
\qquad
\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv}{\overline{\lvalue}}{\overline{\type}}
}{
\expTypJudC{\tctx}{\TStypEnv,\dc: \forall \overline{\tvar}\overline{\bvar}. \type'}{\TtypEnv}{\dc(\overline{\lvalue})}{\type} }
}
%
\skiptransition
%
\surfaceTyping{T-A-AFUN}{ \quad
\expTypJudC{\aann}{\TStypEnv}{\;\TtypEnv,\,\overline{\xname}:\overline{\type}}{\e}{\type}
}{ \expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{ (\overline{\xname}) \; \toSym{\name}{\{\alignK\{\e\}\}}}{(\overline{\type})\rightarrow\type}}
%
\skiptransition
%
\surfaceTyping{T-A-SFUN}{ \quad
\expTypJudC{\sann}{\TStypEnv}{\;\TtypEnv,\,\overline{\xname}:\overline{\type}}{\e}{\type}
}{ \expTypJudC{\sann}{\TStypEnv}{\TtypEnv}{ (\overline{\xname}) \; \ftoSym \{\e\}}{(\overline{\type})\rightarrow\type}}
%
\skiptransition
%
{
\surfaceTyping{T-N-AFUN}{ \quad
\mbox{$\afunvalue$ is a (built-in or declared) aggregate function}}{
\expTypJudC{\aann}{\TStypEnv,\afunvalue: \forall \overline{\tvar}\overline{\bvar}. \type}{\TtypEnv}{\afunvalue}{\applySubstitution{\applySubstitution{\type}{\substitution{\overline\tvar}{\overline\type}}}}{\substitution{\overline\bvar}{\overline\builtintype}} }
}
%
\skiptransition
%
{
\surfaceTyping{T-N-SFUN}{ \quad
\mbox{$\sfunvalue$ is a (built-in or declared) standard function}}{
\expTypJudC{\sann}{\TStypEnv,\sfunvalue: \forall \overline{\tvar}\overline{\bvar}. \type}{\TtypEnv}{\sfunvalue}{\applySubstitution{\applySubstitution{\type}{\substitution{\overline\tvar}{\overline\type}}}}{\substitution{\overline\bvar}{\overline\builtintype}} }
}
%
\skiptransition
%
\surfaceTyping{T-APP}{ \quad
\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv}{\e}{(\overline{\type})\rightarrow\type} \qquad
\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv}{\overline{\e}}{\overline{\type}} }{
\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv}{\e(\overline{\e})}{\type} }
%
\skiptransition
%
%\surfaceTyping{T-BRANCH}{ \qquad
%\expTypJud{\TStypEnv}{\TtypEnv}{\e_1}{\btype}
%\quad \expTypJud{\TStypEnv}{\TtypEnv}{\e_2}{\type}
%\quad \expTypJud{\TStypEnv}{\TtypEnv}{\e_3}{\type} }{
%\expTypJud{\TStypEnv}{\TtypEnv}{\ifK(\e_1) \{\e_2\} \{\e_3\} }{\type} }
%%
%\skiptransition
%%
\surfaceTyping{T-REP}{ \qquad
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\e_1}{\builtintype}
\qquad \expTypJudC{\aann}{\TStypEnv}{\TtypEnv,\xname:\type}{\e_2}{\builtintype} }{
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\repK(\e_1)\{(\xname) \ftoSym \e_2\}}{\builtintype} }
%
\qquad\qquad
%
\surfaceTyping{T-NBR}{ \qquad
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\e}{\builtintype}
}{ \expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\nbrK\{\e\}}{\ftypeOf{\builtintype}} }
%
\skiptransition
%
\surfaceTyping{T-SHARE}{ \qquad
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\e_1}{\builtintype}
\qquad \expTypJudC{\aann}{\TStypEnv}{\TtypEnv,\xname:\ftypeOf{\builtintype}}{\e_2}{\builtintype} }{
\expTypJudC{\aann}{\TStypEnv}{\TtypEnv}{\shareK(\e_1)\{(\xname) \ftoSym \e_2 \}}{\builtintype} }
%
\skiptransition
%
\end{array}
\\
\textbf{Function typing:} 
  \hfill
  \boxed{\funTypJud{\TStypEnv}{\FUNCTION}{\typescheme}}
  \\
\begin{array}{c}
%
\surfaceTyping{T-AFUNCTION}{
\qquad
\expTypJudC{\aann}{\TStypEnv,\,\afname:\forall\emptyseq.(\overline{\type})\rightarrow\type}{\overline{\xname}:\overline{\type}}{\e}{\type}
\qquad
\overline{\tvar}\overline{\bvar}=\FTV{(\overline{\type})\rightarrow\type}
}{ \funTypJud{\TStypEnv}{\funK \; \afname (\overline{\xname}) \; \{ \alignK{\{\e\}} \}}{\forall\overline{\tvar}\overline{\bvar}.(\overline{\type})\rightarrow\type}}
%
\skiptransition
%
%
\surfaceTyping{T-SFUNCTION}{
\qquad
\expTypJudC{\sann}{\TStypEnv,\,\sfname:\forall\emptyseq.(\overline{\type})\rightarrow\type}{\overline{\xname}:\overline{\type}}{\e}{\type}
\qquad
\overline{\tvar}\overline{\bvar}=\FTV{(\overline{\type})\rightarrow\type}
}{ \funTypJud{\TStypEnv}{\funK \; \sfname (\overline{\xname}) \; \{ \e\}}{\forall\overline{\tvar}\overline{\bvar}.(\overline{\type})\rightarrow\type}}
%
\skiptransition
%
\end{array}
\\
\textbf{Program typing:} 
  \hfill
  \boxed{\proTypJud{\PROGRAM}{\type}}
  \\
\begin{array}{c}
%
\surfaceTyping{T-PROGRAM}{
\\
\TStypEnv_0=\OStypEnv
\\
\FUNCTION_i = \funK \; \fname_i (\_) \; \feqSymK{\_}
\qquad
\funTypJud{\TStypEnv_{i-1}}{\FUNCTION_i}{\typescheme_i}
\qquad
\TStypEnv_i=\TStypEnv_{i-1},\, \fname_i:\typescheme_i
\qquad
 (i \in 1..n)
\\
\expTypJudC{\aann}{\TStypEnv_n}{\emptyset}{\e}{\type}
}{ \proTypJud{\FUNCTION_1\cdots\FUNCTION_n  \;
\e}{\type}}
\end{array}
\end{array}$}
} \caption{Type rules for \FKotac{} expressions, function declarations, and programs.} \label{fig:fkotacTyping}
\end{figure}

Figure \ref{fig:fkotacTyping} shows the typing rules for \FKotac{}. A type $\type$ can be a type variable $\tvar$, a built-in type $\builtintype$, a function type $\overline{\type}\to\type$ or a field type $\ftypeOf{\builtintype}$ of a built-in type  $\builtintype$. A built-in type $\builtintype$ is either a built-in variable type $\bvar$ (which actual value can be only a built-in type) or a basic built-in type $\bitype$. Type schemes $\typescheme$ can produce regular types $\type$ by replacing the type variables $\overline{\tvar}$ with types $\overline{\type}$ and the type variables $\overline{\bvar}$ with types $\overline{\builtintype}$. The typing judgement rules are given in the form $\expTypJudC{\tctx}{\TStypEnv}{\TtypEnv}{\e}{\type}$, meaning that $\e$ has type $\type$ under the context $\tctx$, the type-scheme assumptions $\TStypEnv$ and the type assumptions $\TtypEnv$- A context $\tctx$ can be either an aggregate context $\aann$ or a standard context $\sann$, the former meaning that the typing is currently happening inside an aggregate function or expression, the latter inside a standard function or expression.  A type scheme environement $\TStypEnv$ collect the type schemes for built-in constructors, built-in operators and user-defined functions. A type environment $\TtypEnv$ collects the types for program variables. $\OStypEnv$ denotes the distinguished built-in type-scheme environment for built-in constructors and built-in functions. $\FTV{\type}$ denotes the set of type variables occurring in the type $\type$.

Rules \ruleNameSize{[T-ATOS]} and \ruleNameSize{[T-STOA]} allows arbitrary change of context for expressions of built-in types. Those values are the only one allowed to be passed to standard functions to aggregate functions and viceversa. Rule \ruleNameSize{[T-VAR]} assigns in any context a type to a variable $\xname$ by looking the type assumptions $\TtypEnv$. Rule \ruleNameSize{[T-DAT]} assigns in any context a type to a local data value by extracting the type-scheme of the data constructor from the type-scheme environment $\TStypEnv$ and replacing the type variables with types. Rules \ruleNameSize{[T-A-AFUN]} and \ruleNameSize{[T-A-SFUN]} assign a type to an anonymous aggregate function and an anonumous standard function respectively, the former in the aggregate context, the latter in the standard context. Rules \ruleNameSize{[T-N-AFUN]} and \ruleNameSize{[T-N-SFUN]} assign a type to a built-in or declared function, the former in the aggregate context in case of an aggregate function, the latter in the standard context in case of a standard function, both by the function type scheme in the type-scheme assumptions. Rule \ruleNameSize{[T-APP]} assigns a type to a function application, by typing in the same context the function and the arguments. Rule \ruleNameSize{[T-REP]} assigns in the aggregate context a type to a $\mathtt{rep}$-expression, ensuring that the initial value and the body have the same built-in type. Rule \ruleNameSize{[T-NBR]} assings in the aggregate context a field type to a $\mathtt{nbr}$-expression. Rule \ruleNameSize{[T-SHARE]} assigns in the aggregate context a type to a $\mathtt{share}$-expression, enrsuring that the initial value and the body have the same built-in type.

Rules \ruleNameSize{[T-AFUNCTION]} and \ruleNameSize{[T-SFUNCTION]} assigns a type scheme to a user defined function, aggregate and standard respectively, each in the appropriate context. Rule \ruleNameSize{[T-PROGRAM]} allows to type a program with the type of the main expression in the aggregate context by adding to the type-scheme environment the type schemes of all the user defined functions.
